---
title: "lab_02"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

**Q1.** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most? **A1.** 32 departments offer courses with climate in the title, but the Atmospheric and Oceanic Science department gets =

```{r}

climate_source <- umd_courses |> 
  filter (str_detect(title,"Climate") & seats >= 1) #im setting up a new variable that is a filtered data set. 

department_count <- climate_source |> 
  group_by(department) |> 
  summarise(count = n()) |>  # this should count the classes per department unlike the first code block..
  summarise(total_climate_departments = n()) #ohh this summaries gives the new title for data outputs... makes sense 
print(department_count)

department_climate <- climate_source |> 
  group_by(department) |> 
  summarise(total_climate_source= n()) |>
  arrange(desc(total_climate_source)) 
    

head(department_climate)
    
 



```

**Q2.** Of the departments listed in A1, which one is the most interesting to you, and why? **A2.**
Mechanical Engineering, 11 courses seems like a lot but climate and environment affects engineering and physics.  

**Q3** Which departments offered the most courses in the Fall 2024 term? And which departments in that term have at least one syllabus posted? Answer both of those questions and then compare the top five departments - how are they different?

**A3.**
Business and management department had the top result in both questions, but that may be because the department has the most classes overall. I find it interesting that the music department offered one of the most courses in the fall 2024 term and high syllabi, but the history department doesnt have high number of syllabi 
```{r}
department_fall2024 <- umd_courses |> 
  filter(term == 202408) |> #the rest of the data that would be output woukd be fromthis filter 
  group_by(department) |> 
  summarise(
    total_dep_f2024= n() )|>
  arrange(desc(total_dep_f2024)) 

print(department_fall2024)

department_syll <- umd_courses |>
  filter(syllabus_count >= 1 & term == 202408 ) |>
  group_by(department) |> 
  summarise(total_syll= n()) |>
  arrange(desc(total_syll)) 

print(department_syll)


```

**Q4.** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr, saving the result as a new dataframe. Then create a new column for the month based on expenditure_date. Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name, and we're not looking in payee_name). Write a sentence describing the pattern of spending over time. 
**A4.** The spending increased over the span of months and by a lot from the end of the year campaign (october). 

```{r}

maryland_expenses <- maryland_expenses |> 
  mutate(expenditure_date = as.character(expenditure_date)) 

library(lubridate)
maryland_expenses <- maryland_expenses |> 
  mutate(expenditure_date = mdy(expenditure_date))


maryland_expenses |>
  mutate(month = floor_date(expenditure_date)) |>
  group_by(month) |> 
  summarise(total_amount = sum(amount, na.rm = TRUE)) |> 
  arrange(desc(total_amount))

wes_moore_expenses <- maryland_expenses|>
filter(str_detect(committee_name, "Moore")) |>
mutate(month = floor_date(expenditure_date, "month")) |>
  group_by(month) |>
  summarise(total_amount = sum(amount, na.rm = TRUE)) |>
  arrange(desc(total_amount))

print(ses_moore_expenses)
```

**Q5.** Using case_when(), create a column in the Maryland expenses data called `spending_location` indicating whether each record spent money in Maryland or outside Maryland. For records that are in Maryland, make the new column's value "In-state" and for all others, make it "Out of state". Then write code that shows the total amount of money spent in each category and describe the results. You can do this in one statement or break it up into two statements (saving the new column to your dataframe). **A5**

```{r}
ses_moore_expenses <- maryland_expenses |> 
  filter(str_detect(committee_name, "Moore")) |>
mutate(
  spending_location = case_when(
    str_detect(address, "Maryland") ~ "In-State",
    .default = "Out of State") #.default 
) 

spending_summary <- ses_moore_expenses |>
  group_by(spending_location) |>
  summarise(
    total_spent = sum(amount)
  )

spending_summary
```

**Q6.** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab. **A6.** percentage of moore spending locations. I still want to know if it can show like percent sign or not, would that hav to deal with format

```{r}
ses_moore_expenses <- maryland_expenses |> 
  filter(str_detect(committee_name, "Moore")) |>
  mutate(
    spending_location = case_when(
      str_detect(address, "Maryland") ~ "In-State",
      TRUE ~ "Out of State")
  )

spending_summary <- ses_moore_expenses |>
  group_by(spending_location) |>
  summarise(
    total_spent = sum(amount, na.rm = TRUE)
  ) 

total_spending <- sum(spending_summary$total_spent) #calculate soending for all the locations

spending_summary <- spending_summary |>
  mutate(percentage = (total_spent / total_spending) * 100) # i wanted to use the mutate to make it a percentage instead of dates. 

print(spending_summary)

```
