---
title: "lab_07"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   A Census API key 

## Load libraries and establish settings

You'll need to load two packages for this: the tidyverse and tidycensus.

**Task** load these two packages

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
library(janitor)
library(lubridate)
library(tidycensus)
```

## Setup Census API

You'll need your Census API Key:

**Task** set your API Key if you haven't already. You won't need to install or overwrite it.

```{r echo=FALSE}
census_api_key("c835b290fcfd849bd146850c22c7b7ddda145ae2")
```

## Load and view ACS variables

You'll also need to load the list of American Community Survey variables from the 2023 5-year estimates so you can identify the codes you need:

**Task** load the variables from the 2023 5-year ACS (which is called "acs5") and save them to a dataframe

```{r}
acs5 <- load_variables(2023, "acs5", cache = TRUE)

head(acs5)
```

## Answer questions

**Q1** What is the Maryland jurisdiction with the lowest median age, according to the ACS 2023 5-year estimates? You will need to find the variable for median age first. **A1** Baltimore City, Maryland has the lowest median age, with an estimate of 36.1 

```{r}

#call it maryland. to find variables, use the loading variable thing to find the different 
md_median <- get_acs(geography = "county",
              variables = c(mediage = "B01002_001"),
              state = "MD",
              year = 2023)

 md_median |> arrange(estimate) |> top_n(24, estimate)
  
  
```

**Q2** Which Maryland counties have a median age of at least 50? You can use the dataframe you just made to answer this, but you must write code to display only those counties with a median age of at least 50. **A2** Talbot County and Worcester County. 

```{r}
md_median|>
  filter (estimate >=50)
  
```

**Q3** We're interested in knowing more about Montgomery County zip codes where overdose calls are coming from and the demographic characteristics they possess. In particular, we want to calculate a rate of calls per 1,000 population. To do this, you'll need to:

1.  Load the Montgomery County 911 overdose calls data.
2.  Create a new dataframe with the total number of calls for each zip code.
3.  Find the ACS variable pertaining to the total 18 and up population (there are several choices here; you want the simplest one).
4.  Get from the ACS data from all Maryland zip codes using that variable - you will need to specify an earlier year to get the data.
5.  Join the ACS data to the Montgomery County zip code totals dataframe and rename the estimate column to a meaningful name you've given the ACS variable.
6.  Add a column to the joined dataframe that calculates a rate of calls per 1,000 population.

Which zip code has the highest rate of calls? Which has the lowest?
20889 had the most calls while 20707 had the least
**A3**

```{r}
moco_overdoses <- read_csv("data/montgomery_2022_overdoseses.csv")

#make a dataset with the totalnumber of calls for each zipcode

moco_zipcode_calls <- moco_overdoses |> group_by(zip) |> summarise(calls = n())

moco_zipcode_calls 

#then i need to find the acs variable that has total 18 and up population: ->  B09021_001 Estimate!!Total: Living Arrangements of Adults 18 Years and Over

#zcta means zip code remember, get acs data from all md zipcodes using that variable 

over_18_md <- get_acs(geography = "zcta",
                           variables = "B09021_001",
                           year = 2022)

#join the acs data to moco zip code totals data frame and rename the estimate column to a meaningful name you've given the acs variables 

over_18_md_2022 <- over_18_md |>
  clean_names() |>
  rename(
    zip = geoid
  )
 
moco_zipcode_calls <- moco_zipcode_calls |>
  mutate(
    zip = as.character(zip)
  )

moco_zipcode_calls <-moco_zipcode_calls |>
  left_join(over_18_md_2022) |>
  rename(over18 = estimate)
#adding column to joun data frame that calcuate rate of calls per 1000pop
moco_zipcode_calls <- moco_zipcode_calls|>
  mutate(
    calls_pct = calls/over18 * 1000
  )
                           
moco_zipcode_calls |>
  arrange(desc(calls_pct))
```


**Q4** Using [CensusReporter.org](https://censusreporter.org/), examine the zip codes with the highest and lowest rates and compare and contrast their demographic profiles. Is there a story here?

**A4** Zipcode areas with high overdose rates have mostly white men with 100k plus salary, which contrast from the areas with low rates of overdoses have diverse demographics. There could be a story about how certain populations, in this case white men with high finances report and call for overdoses more than diverse demographics. Maybe there is more resources in those zipcodes 

**Q5** Go back to your joined dataframe from Q3 and add two more columns representing the minimum and maximum populations based on the estimate and margin of error, then calculate per capita rates for each. Do you get the same two zip codes for the highest and lowest as you did before? Does that give you more or less confidence in your answers to Q3?

**A5** same zipcodes as before for the max population but the minimum population are different zipcodes. Whenb a population is lower than the margin of error(like 20988), there will be negative minimum population and holds each overdose call with a higher weight to per capita value compared to other zipcodes (the ratio is smaller)

```{r}
moco_zipcode_calls <- moco_zipcode_calls |>
  mutate(
    min_popu = over18 - moe, #margin of error,
    max_popu = over18 + moe
)

moco_zipcode_calls <- moco_zipcode_calls |>
  mutate(
    min_pct = calls/min_popu *1000, 
    max_pct = calls/max_popu *1000
  )
moco_zipcode_calls
```
