---
title: "lab_04"
author: "Derek Willis"
date: "2023-10-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load three packages for this: the tidyverse, lubridate and janitor.

**Task** load these three packages.

```{r}
# Turn off scientific notation
options(scipen=999)

# load your packages
library(tidyverse)
library(lubridate)
library(refinr)

```

For this lab, we want to investigate spending by Maryland state and local candidates during the 2022 election. For example, we could ask how much money have candidates and committees spent on advertising on Facebook? Which vendors received the most money for media expenses? We have the data, but it's messy - names of payees are spelled differently and capitalized differently - and we need to clean it up. We'll use a combination of RStudio and OpenRefine to do that.

The basic process here is to start in RStudio, export data so that we can use OpenRefine to clean it up, and then bring it back into RStudio to finish our analysis.

## Load Data

You'll need to load one data set, a CSV file of Maryland campaign expenditures located in the data folder called "maryland_expenses.csv"

**Task** Create a codeblock below, then read the data in and assign it to an appropriate variable name. You'll want to clean up the column names and make sure that any date columns are actually date datatypes.

```{r}
# Load the Maryland expenditures data table

library(readr)
maryland_expenses <- read_csv("Documents/GitHub/data_journalism_2025_spring/labs/lab_04/data/maryland_expenses.csv")
(maryland_expenses)


library(dplyr)      # For data manipulation
library(lubridate)  # For date handling
library(stringr)    # For string manipulation
library(janitor)    # For cleaning column names

# Display the cleaned data frame
cleaned_md_expenses <- maryland_expenses |> 
  mutate(date = mdy(`Expenditure Date`)) |> 
  clean_names() |> 
  mutate(
    payee_name = str_to_upper(payee_name),
    committee_name = str_to_upper(committee_name),
    description = str_to_upper(comments),
    expense_method = str_to_lower(expense_method),
    
  )




 

head(cleaned_md_expenses)



```

## Answer questions

**Q1.** You've been assigned to write about Maryland political committees' use of Facebook advertising. There's a category column in the data that has values for different kinds of spending. One of them is "Media". You'll need to complete the following steps:

1.  Create a dataframe of Media expenses where the name of the payee is not NA.
2.  Then, using `write_csv`, make a CSV file of your media expenses in your data folder. `write_csv()` is similar to `read_csv()` but you provide it with the dataframe you're writing and the file path.
3.  Using OpenRefine, create a project using that CSV file, create a copy of the `payee_name` column called `payee_clean` and then standardize it, focusing on the payees that appear most often. Pay particular attention to variations of Facebook, bearing in mind that it changed its name in 2021: <https://about.fb.com/news/2021/10/facebook-company-is-now-meta/>), but also clean up variations of the most common payees.
4.  When you are done standardizing `payee_clean`, export it as a new CSV file to your data folder, *giving it a different name* and read it back into RStudio as a new dataframe.
5.  Using that `payee_clean` column in your new dataframe, calculate the total amount of Facebook spending by each committee and present the results showing the committee that spent the most first.

**A1.**

```{r}
mrmedia <- cleaned_md_expenses |>
  filter(
    expense_category == "Media", 
    payee_name != "NA"
    )

head(mrmedia)

#make a csv file for media expenses file path

write_csv(mrmedia, "data/mrmedia")

#after cleaning on Open Refine 

clean_mrmedia <- read_csv("data/clean_mrmedia.csv")

head(clean_mrmedia)
# calculate the total amount (at the end, summarize and the sum of  amount) of Facebook (or meta) spending by each committee (group by committee) and present the results showing the committee that spent the most first (arrange descendign order) .

clean_mrmedia |>
  filter(str_detect(payee_clean, "META")) |>
  group_by(committee_name) |>
  summarize(total_mrspend =sum(amount)) |>
  arrange(desc(total_mrspend))

```

**Q2.** Let's do some research on the top committees from question 1. Look up the top 10 spenders: if they are candidate committees, what offices were they running for? If they were not candidates, what issues do they seem to care about? Write some sentences answering those questions, then go the Facebook ad library ([https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data\\[direction\\]=desc&sort_data\\[mode\\]=relevancy_monthly_grouped&media_type=all](https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data%5Bdirection%5D=desc&sort_data%5Bmode%5D=relevancy_monthly_grouped&media_type=all){.uri}) and look for ads from the top spender. What do you notice about the the information provided about those ads?

**A2.**
Candidates for office --
- JASANI SALIMAH : Baltimore City Board of School Commissioners
- SHUMAN EVE T. : Democratic Candidate for PG County Council District 3
- MAX SOCOL : Maryland Senate
- MINK KRISTIN, BROOKS BRANDY, NATALI FANI-GONZALEZ : Montgomery County Council
- JESSE PIPPY : Maryland House of Delegate (Frederick)
- RICHARD D FRITZ : Maryland State Attorney (St. Mary)
- Izzy Pakota - Baltimore County Council. 
Of those that were not candidates for office --
- CASA is an organization that advocates for immigrant rights.

When looking at the one of the spenders ad dat, it shares the estimated audience size, amount spent in USD, and the ads impressions. Theyre in ranges though. CASA spend under 100 dollars for each off its 300+ ads 

**Q3.** Given the list of top spenders in A1 and the primary results for governor (<https://elections.maryland.gov/elections/2022/primary_results/gen_results_2022_1.html>), who is missing from that Q1 list? Did those campaigns run Facebook ads? Who are the biggest recipients of media expenses by the Democratic primary winner for governor (HINT: that person has two committees, the second one with his running mate)?

**A3.** my first thoughts of who is missing from the Meta's ad data is Dan Cox and Wes Moore, the primary candidates for Maryland elections. However when I look at the current governor Moore's, he ran adds during 2021-2022. Usually paid by Mary Tyding treasure. #after the coding, the biggest media expenses by Wes Moore's committee are SKDK, a political consulting firm; Foglamp, an advertising firm; and Apollo Artistry, a digital developer. 


```{r}
clean_mrmedia |>
  filter(committee_name == "MOORE  WES FOR MARYLAND" | committee_name == "MOORE MILLER FOR MARYLAND SLATE") |>
  group_by(payee_clean) |>
  summarise(total_spend_moore = sum(amount)) |>
  arrange(desc(total_spend_moore))

```

**Q4.** The media expense data you cleaned also has a standardized purpose column. Write code to generate a list of the committees with the total amount of spending on online advertising. How does this list compare to the one from Q1 about top Facebook spenders? Of the top 5 committees who spent the most on Facebook ads from Q1, are there any that appear to have spent *all* of their online advertising dollars on Facebook?

**A4.**
CASA  seems to also appear on the list of committees that spend making online advertising media ads/ It looks like the committees for Ashwani Jain spent all of their online advertising money on Facebook.

```{r}
clean_mrmedia |>
  filter(expense_purpose == "Online Advertising")|>
  group_by(committee_name) |>
  summarise(total_spend_ad = sum(amount)) |>
  arrange(desc(total_spend_ad))
  
```

**Q5** Your editor asks you to summarize your findings about spending on Facebook by Maryland political candidates. How much do we know and what don't we know? What's the best story idea you've found here?

**A5**
We see the candiates and how much they spent within Meta/Facebook, but we cant differiate if the ads are on Instagram or just Facebook. The dataset by itself doesn't soley show the type of ad, that has to be combined with the Meta ad library. With that, I can figure out which type of parties that are spending the most on Facebook ads for the 2022 maryland expenses. But For know I can figure out the total spending of all the spending on Meta. I put all the facebook and meta as just Meta.     

```{r}
cleaned_maryland_expenses_media |> 
  filter(str_detect(payee_clean, "Meta")) |>
  summarise(total_spend = sum(amount))

