---
title: "lab_10"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Our usual libraries, plus some of the map-specific ones from the pre_lab. You'll also be fetching Census data in Q3. Read the questions carefully!

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
# Load libraries
library(tidyverse)
library(sf)
library(janitor)
library(maptools)
library(tigris)
```


Let's explore Maryland elections, power outages and electric vehicle data and make some maps!

## Questions

**Q1.** Maryland voters made history last November by electing the state's first two African-American statewide office holders: Wes Moore as governor and Anthony Brown as attorney general. You want to compare the two's electoral performances. Using the `md_statewide_county.csv` file in the data folder, 

- calculate the total number of votes for both the governor's race and the attorney general's race, 
- then percentages for both Moore and Brown *based on the total of their respective contests*.
- Then calculate the difference between their percentages.

Get the geometry for Maryland's counties using tigris and, joining that dataframe to your results dataframe, make a county-level map of the differences between Moore and Brown's percentages. What is the story here, and what is the county with the most interesting results?

**A1.** Kent and Carroll had the highest differences between Moore election votes and Brown. Since they had higher difference in those state. But the states with higher differences tend to be more Red states so its intresting  

```{r}
md_statewide_county <- read_csv("data/md_statewide_county.csv")|>
  clean_names()

md_statewide_county <- md_statewide_county|>
  mutate(
    total_gov_vote = cox + moore + lashar + wallace + harding + sefcik + gov_writein,
    total_ag_vote = peroutka + brown + ag_writein,
    pct_moore = moore /total_gov_vote *100,
    pct_brown = brown / total_ag_vote *100,
    ag_gov_diff = abs(pct_moore - pct_brown)
  )
electoral_vote <- md_statewide_county |>
  mutate(
     geoid = as.character(geoid)
  )


#Geometry for Marylands counties 

md_counties <- counties |>
  filter(STATEFP == "24")|>
  clean_names()

md_counties
```



```{r}
election_map_county <- md_counties |> left_join(electoral_vote)

#joining dataframes to map

ggplot() +
  geom_sf(data=election_map_county, aes(fill=ag_gov_diff)) +
  scale_colour_viridis_b(option="magma") + 
  geom_text(aes(x = X, y = Y, label = NAME), data = county_centroids_df, size = 3, check_overlap = TRUE) +
  theme_minimal() 
```

**Q2.** Make a county-level map to explore the number of customers impacted by power outages in Maryland from last March. Load the `county_outages_march2023.csv` file in the data folder and 
calculate a rate of customers with outages per 10,000 total customers for each county. 
Then join that dataframe to your Maryland counties dataframe you made in Q1 and create a county-level map of the results; you may need to tweak the fill scale to make the distinctions clearer. What's the potential story here? Given access to other months' outage data, what would your next steps be in reporting a story?

**A2.**  Talbot County has the most number of outages for any county for March 2023. Another story could be that the counties that are far from the Chesapeake Bay, seem to have less outages. If I had access to other months' data, I would look to see if these patterns hold true over time if Tablot is an outlier 

```{r}
#load data

county_outages_march2023 <- read_csv("data/county_outages_march2023.csv")

head(county_outages_march2023)


 outages3_23 <- county_outages_march2023|>
  mutate(
   outage_per_10k = outage_customers / total_customers * 10000
  )

outages3_23 <- outages3_23 |>
  rename(name = county)

outages_map <- md_counties |>
  left_join(outages3_23)

#map of the results
ggplot() +
geom_sf(data=outages_map, aes(fill=outage_per_10k)) +
scale_colour_viridis_b(option="magma", trans = "log") + 
  geom_sf_label(data=outages_map, aes(label=name), 
  nudge_x = 0.15, nudge_y = 0.0, check_overlap = FALSE, size = 2.5, color = "black", fill = "white" )#how do i add text to the counies with a padding border?? its not working (notes fo rfuture )
theme_minimal()




```

**Q3.** You've been assigned a story about the use of electric vehicles in Maryland and you want to make a map of the prevalence of electric vehicles in each county. Load the `ev_2020_2023.csv` file, which has the number of electric and hybrid vehicles per county from July 2020 and March 2023 and calculate a percentage change for each county (remember, that's `new-old/old`). Using the Maryland counties dataframe you made for Q1, make a county map showing the percentage change. What counties stand out to you? What questions do you have about the changes you're seeing?

Next, use tidycensus to retrieve county population data for Maryland from 2021 (you did this in the pre_lab) and join it to your vehicles dataframe. HINT: you may not need to join it on the county name, and you should already have the geometry!

Then, calculate a per capita rate of electric & hybrid vehicles using the March 2023 figure and the population estimate. You will need to choose the "per capita" value.

Finally, make a second county map showing the per capita rate. What counties stand out on this map? How is it different from the first one you made?

**A3.**  In the first map, it is Caroline that shows the most percentage change for non gas motor vehicles, but when finding the rate with the per capia, the second map reveals that Howard and Montgomery counties has the most usage of electric or hybrid vehicles. With Caroline and Somerset, they have smaller numbers so the precent change will be affected stronger than counties with larger numbers of vechiles already. But I wonder why those counties have gotten more increase of electric vechile sover the years. 

For the second map,most of the counties that have a low percent change tended to have a higher per capita rate. They are all clustered together near the center fo maryland, but also since the popualtion is aalready high, the would cuase more electric vehicles so the percent change would barelty be affected. 


```{r}
# Load data
electric_vehicles <- read_csv("data/ev_2020_2023.csv")

# Calculate a percentage change for each county
electric_vehicles <- electric_vehicles |>
  mutate(
    pct_change = (march_2023 - july_2020) / july_2020 * 100
  )

# make a county map showing the percentage change.
electric_vehicles <- electric_vehicles |>
  rename(name = county)
  
elec_vehic_geog <- md_counties |>
  left_join(electric_vehicles)

ggplot() +
  geom_sf(data=elec_vehic_geog, aes(fill=pct_change)) +
  scale_colour_viridis_b(option="magma", trans = "log") + geom_sf_label(data=elec_vehic_geog, aes(label=name), 
                nudge_x = 0.25, nudge_y = 0.0, check_overlap = FALSE, size = 2.5, color = "black", fill = "white" ) +
  theme_minimal()


# retrieve county population data for Maryland from 2021
md_county_popu <- get_acs(geography = "county",
              variables = c(population = "B01001_001"),
              year = 2021,
              state = "MD") |>
  clean_names() |>
  select(geoid, estimate) |>
  rename(population = estimate)

# Join it to your vehicles dataframe
ev_geog_popu <- elec_vehic_geog |>
              select(geoid, name, march_2023, geometry) |>
  left_join(md_county_popu)

# Calculate a per capita rate of electric & hybrid vehicles using the March 2023 figure and the population estimate
ev_geog_popu <- ev_geog_popu |>
  mutate(
    per_cap = march_2023 / population * 1000
  )

# Make a second county map showing the per capita rate
ggplot() +
  geom_sf(data=ev_geog_popu, aes(fill=per_cap)) +
  scale_fill_viridis_c(option = "magma", trans = "log") +  
  geom_sf_label(data=ev_geog_popu, aes(label=name), 
                nudge_x = 0.25, nudge_y = 0.0, check_overlap = FALSE, size = 2.5, color = "black", fill = "white" ) +
  theme_minimal()


```
